---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# motif

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/Nowosad/motif.svg?branch=master)](https://travis-ci.org/Nowosad/motif)
[![GitHub action build status](https://github.com/Nowosad/motif/workflows/pkgdown/badge.svg)](https://github.com/Nowosad/motif/actions)
[![Codecov test coverage](https://codecov.io/gh/Nowosad/motif/branch/master/graph/badge.svg)](https://codecov.io/gh/Nowosad/motif?branch=master)
<!-- badges: end -->

The **motif** package implements ideas of the pattern-based spatial analysis in R.
It describes spatial patterns of categorical raster data for any defined regular and irregular areas. 
Patterns are represented quantitatively using built-in signatures based on co-occurrence matrices but also allows for any user-defined functions. 
It enables spatial analysis such as search, change detection, and clustering to be performed on spatial patterns.

## Installation

<!-- You can install the released version of motif from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("motif") -->
<!-- ``` -->

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("Nowosad/motif")
```

## Example

### Signatures

```{r}
library(motif)
library(stars)
landcover = read_stars(system.file("raster/landcover2015.tif", package = "motif"))
landcover = droplevels(landcover)
plot(landcover, key.pos = 4, key.width = lcm(5))
```

```{r}
landcover_coma = lsp_thumbprint(landcover, type = "coma", window = 200)
landcover_coma
```

```{r}
landcover_coma$signature[[1]]
```

<!-- ref to vig -->
<!-- create a summary image -->
<!-- ### Search -->
<!-- ### Compare -->
<!-- ### Cluster -->

## Contribution

Contributions to this package are welcome.
The preferred method of contribution is through a GitHub pull request. 
Feel free to contact us by creating [an issue](https://github.com/Nowosad/motif/issues). 
