---
title: "Spatial patterns' clustering"
author: "Jakub Nowosad"
date: "`r Sys.Date()`"
output: 
  html_document:
  github_document:
    html_preview: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

The pattern-based spatial analysis makes it possible to find clusters of areas with similar spatial patterns.
This vignette shows how to do spatial patterns' clustering on example datasets.
Let's start by attaching neccesary packages:

```{r}
library(motif)
library(stars)
library(sf)
safe_pal6 = c("#88CCEE", "#CC6677", "#DDCC77",
             "#117733", "#332288", "#888888")
safe_pal4 = c("#88CCEE", "#CC6677", "#DDCC77", "#117733")
```

```{r, echo=FALSE, eval=FALSE}
# dput(rcartocolor::carto_pal(n=6, "Safe"))
```

Spatial patterns' search clustering having one spatial object.
For this vignette, we will read the `"raster/landcover2015.tif"` file.

```{r}
landcover = read_stars(system.file("raster/landcover2015.tif", package = "motif"))
```

This file contains a land cover data for New Guinea, with seven possible categories: (1) agriculture, (2) forest, (3) grassland, (5) settlement, (6) shrubland, (7) sparse vegetation, and (9) water.

```{r}
landcover = droplevels(landcover)
plot(landcover, key.pos = 4, key.width = lcm(5), main = NULL)
```

## Regular local landscapes

In the first example, we will divide the whole area into many regular local landscapes, and find a way to cluster them based on their patterns.

```{r, echo=FALSE}
stars200 = lsp_add_sf(landcover, window = 100)
landcover = droplevels(landcover)
plot(landcover, key.pos = NULL, reset = FALSE, main = NULL)
plot(st_geometry(stars200), add = TRUE, reset = FALSE)
# st_bbox(stars200)
# st_bbox(landcover)
```

It starts with calculating a signature for each local landscape using the `lsp_thumbprint()` function.
Here, we will use the co-occurrence vector (`type = "cove"`) representation on local landscapes of 100 by 100 cells (`window = 100`).
Distance measures, used in one of the next steps, often require normalization of the vector, therefore we transform the representation to sum to one.

```{r}
landcover_cove = lsp_thumbprint(landcover, type = "cove",
                                window = 100, normalization = "pdf")
```

Read the [Introduction to the motif package](https://nowosad.github.io/motif/articles/intro.html) vignette to learn more about spatial signatures.

Next, we need to calculate the distance between each local landscape.
This can be done with the `lsp_to_dist()` function, where we must provide the output of `lsp_thumbprint()` and a distance/dissimilarity measure used (`dist_fun = "jensen-shannon"`).

```{r}
landcover_dist = lsp_to_dist(landcover_cove, dist_fun = "jensen-shannon")
```

The output, `landcover_dist`, is of a `dist` class.

```{r}
class(landcover_dist)
```

This allows it to be used by many existing R functions for clustering, which expect a distance matrix (`dist`) as an input.
It includes different approaches to hierarchical clustering (`hclust()`, `cluster::agnes()`, `cluster::diana()`), fuzzy clustering (`funny()`).
More information about clustering techniques in R can be found in [CRAN Task View: Cluster Analysis & Finite Mixture Models](https://cran.r-project.org/web/views/Cluster.html).

We will use a hierarchical clustering using `hclust()` in this example.
It needs a distance matrix as a first argument and a linkage method as a second one.
Here, we will use `method = "ward.D2"`, which means that we will be using Ward's minimum variance method that minimizes the total within-cluster variance. 

```{r}
landcover_hclust = hclust(landcover_dist, method = "ward.D2")
plot(landcover_hclust)
```

Graphical representation of the hierarchical clustering is called a dendrogram.
Based on the obtained dendrogram, we can divide our local landscapes into a specified number of groups using `cutree()`.
In this example, we will use six classes (`k = 6`) to make a result fairly straightforward to understand.

```{r}
clusters = cutree(landcover_hclust, k = 6)
```

The **motif** package allows representing the clustering results as either `stars` or `sf` object. 
This can be done with the `lsp_add_clusters()` function.

### `sf`

By the default, it creates a `sf` object with a new variable `clust`:

```{r}
landcover_grid_sf = lsp_add_clusters(landcover_cove, clusters)
plot(landcover_grid_sf["clust"], pal = safe_pal6)
```

We can also calculate the quality of the clusterings with the `lsp_add_quality()` function, which requires an output of `lsp_add_clusters` as the first argument and output of `lsp_to_dist()` as the second one.
It adds three new variables: `inhomogeneity`, `isolation`, and `quality`.

```{r}
landcover_grid_sfq = lsp_add_quality(landcover_grid_sf, landcover_dist)
```

Inhomogeneity measures a degree of mutual dissimilarity between all objects in a cluster. 
This value is between 0 and 1, where the small value indicates that all objects in the cluster represent consistent patterns so the cluster is pattern-homogeneous. 

```{r}
inhomogeneity_pal = hcl.colors(6, palette = "Emrld")
plot(landcover_grid_sfq["inhomogeneity"], pal = inhomogeneity_pal)
```

Isolation is an average distance between the focus cluster and all of its neighbors. 
This value is between 0 and 1, where the large value indicates that the cluster stands out from its surroundings.

```{r}
isolation_pal = hcl.colors(6, palette = "Teal", rev = TRUE)
plot(landcover_grid_sfq["isolation"], pal = isolation_pal)
```

Quality is calculated as `1 - (inhomogeneity / isolation)`.
This value is also between 0 and 1, where increased values indicate increased quality. 

```{r}
quality_pal = hcl.colors(6, palette = "Temps", rev = TRUE)
plot(landcover_grid_sfq["quality"], pal = quality_pal)
```


### `stars`

To create an `stars` object with a new attribute `clust`, we should set `output = "stars"`.

```{r}
landcover_grid_stars = lsp_add_clusters(landcover_cove, clusters, output = "stars")
plot(landcover_grid_stars["clust"], col = safe_pal6)
```

<!-- ADD EXAMPLES OF EACH CLASS!! -->
However, it is not yet possible to calculate the quality of the clusterings for the `stars` objects:

```{r, error=TRUE}
landcover_grid_starsq = lsp_add_quality(landcover_grid_stars, landcover_dist)
```

```{r, echo=FALSE, eval=FALSE}
plot(landcover_grid_starsq["inhomogeneity"], col = inhomogeneity_pal)
```
```{r, echo=FALSE, eval=FALSE}
plot(landcover_grid_starsq["isolation"], col = isolation_pal)
```
```{r, echo=FALSE, eval=FALSE}
plot(landcover_grid_starsq["quality"], col = quality_pal)
```

## Irregular local landscapes

The **motif** package also allows for the clustering of irregular regions based on the user-provided polygons.
It has an example spatial vector dataset, `ecoregions.gpkg`, which contains terrestrial ecoregions for New Guinea from https://ecoregions2017.appspot.com/. 

```{r}
ecoregions = read_sf(system.file("vector/ecoregions.gpkg", package = "motif"))
```

This dataset has 22 rows, where each row relates to one ecoregion.
Each ecoregion is also related to a unique value in the `id` column.

```{r, echo=FALSE}
# https://medialab.github.io/iwanthue/
my_pal = c("#d34359", "#62b93c", "#b75fcf", "#53c069", "#d44295", "#acb939", "#626edd", "#dc9e36", "#8156a8", "#4b8734", "#d98dc7", "#58c096", "#cc542a", "#48bbd2", "#bf814d", "#6686c8", "#968c30", "#a34d78", "#36815b", "#c26963", "#a2b36b", "#6b6829")
plot(ecoregions["id"], main = NULL, col = my_pal, key.pos = 4, key.width = lcm(5))
```

The first step here is to calculate a signature for each ecoregion using `lsp_thumbprint()`:

```{r}
landcover_cove2 = lsp_thumbprint(landcover, type = "cove",
                                 window = ecoregions["id"], normalization = "pdf")
```

Next step involves calculating distances between signatures of ecoregions:

```{r}
landcover_dist2 = lsp_to_dist(landcover_cove2, dist_fun = "jensen-shannon")
```

Again, we will find clusters using a hierarchical clustering:

```{r}
landcover_hclust2 = hclust(landcover_dist2, method = "ward.D2")
plot(landcover_hclust2)
```

This time, we will divide the data into four clusters:

```{r}
clusters2 = cutree(landcover_hclust2, k = 4)
```

### sf

Next steps are to add clusters ids to the `sf` object:

```{r}
landcover_grid_sf2 = lsp_add_clusters(landcover_cove2, clusters2,
                                      window = ecoregions["id"])
```

The results show that most area of New Guinea belongs to the second clusters, with only one ecoregion of cluster 3 (small islands south from New Guinea) and cluster 4.

```{r}
plot(landcover_grid_sf2["clust"], pal = safe_pal4)
```

Final step involves of calculating the quality metrics of the obtained clusters:

```{r}
landcover_grid_sfq2 = lsp_add_quality(landcover_grid_sf2, landcover_dist2)
```

```{r}
plot(landcover_grid_sfq2["inhomogeneity"], pal = inhomogeneity_pal[c(1, 3, 6)])
```

```{r}
plot(landcover_grid_sfq2["isolation"], pal = isolation_pal[c(1, 3, 6)])
```

```{r}
plot(landcover_grid_sfq2["quality"], pal = quality_pal[c(1, 3, 6)])
```

### stars

To create an `stars` object with a new attribute `clust`, we should set `output = "stars"`.
However, it is not yet possible to calculate the quality of the clusterings for the `stars` objects.

```{r}
landcover_grid_stars2 = lsp_add_clusters(landcover_cove2, clusters2, 
                                         window = ecoregions["id"], output = "stars")
plot(landcover_grid_stars2["clust"], col = safe_pal4)
```

```{r, echo=FALSE, eval=FALSE}
# stars -------------------------------------------------------------------

landcover_grid_starsq2 = lsp_add_quality(landcover_grid_stars2, landcover_dist2)

plot(landcover_grid_starsq2["inhomogeneity"], col = inhomogeneity_pal)
plot(landcover_grid_starsq2["isolation"], col = isolation_pal)
plot(landcover_grid_starsq2["quality"], col = quality_pal)
```


