---
title: "Spatial patterns' comparision"
author: "Jakub Nowosad"
date: "`r Sys.Date()`"
output: 
  html_document:
  github_document:
    html_preview: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

The pattern-based spatial analysis makes it possible to compare spatial patterns between two datasets.
This vignette shows how to do spatial patterns' comparison on example datasets.
Let's start by attaching neccesary packages:

```{r}
library(motif)
library(stars)
library(sf)
library(tmap)
```

Spatial patterns' comparison requires two spatial objects.
For this vignette, we will read the `"raster/landcover2001.tif"` and `"raster/landcover2015.tif"` files.

```{r}
landcoverold = read_stars(system.file("raster/landcover2001.tif", package = "motif"))
landcover = read_stars(system.file("raster/landcover2015.tif", package = "motif"))
```

The object created based on the first file, `landcoverold` is a land cover of New Guinea in 2001.

```{r}
landcoverold = droplevels(landcoverold)
plot(landcoverold, key.pos = 4, key.width = lcm(5), main = NULL)
```

The second object, `landcover` is a land cover of New Guinea in 2015.

```{r}
landcover = droplevels(landcover)
plot(landcover, key.pos = 4, key.width = lcm(5), main = NULL)
```

Our goal here is to find areas with the largest changes in spatial patterns of land cover.

## Regular local landscapes

Spatial patterns' comparision can be done using the `lsp_compare()` function.
It expects two spatial objects, a type of spatial signature used (`"cove"`), distance function to measure dissimilarity (change) between spatial patterns (`"dist_fun"`).
By default, the whole area is compared, it can be changed with the `window` argument.
Here, `window = 100` means that the study area will be divided into local landscapes of 100 by 100 cells.

```{r}
compare_1 = lsp_compare(landcover, landcoverold,
                        type = "cove", dist_fun = "jensen-shannon",
                        window = 100, threshold = 0.9)
```

The output is a `stars` object with a `dist` argument, with the value of 0 indicating no change, and larger values representing changes of a bigger magnitude.

```{r}
my_breaks = c(0, 0.001, 0.01, 0.1, 1.01)
tm_shape(compare_1) +
  tm_raster("dist", breaks = my_breaks, palette = "-viridis") +
  tm_layout(legend.outside = TRUE)
```

<!-- ADDS EXAMPLES -->

## Irregular local landscapes

The **motif** package also allows for the comparing spatial patterns in irregular regions based on the user-provided polygons.
It has an example spatial vector dataset, `ecoregions.gpkg`, which contains terrestrial ecoregions for New Guinea from https://ecoregions2017.appspot.com/. 

```{r}
ecoregions = read_sf(system.file("vector/ecoregions.gpkg", package = "motif"))
```

This dataset has 22 rows, where each row relates to one ecoregion.
Each ecoregion is also related to a unique value in the `id` column.

```{r, echo=FALSE}
# ecoregions = st_transform(ecoregions, st_crs(landcover))
```

```{r, echo=FALSE}
# https://medialab.github.io/iwanthue/
my_pal = c("#d34359", "#62b93c", "#b75fcf", "#53c069", "#d44295", "#acb939", "#626edd", "#dc9e36", "#8156a8", "#4b8734", "#d98dc7", "#58c096", "#cc542a", "#48bbd2", "#bf814d", "#6686c8", "#968c30", "#a34d78", "#36815b", "#c26963", "#a2b36b", "#6b6829")
plot(ecoregions["id"], main = NULL, col = my_pal, key.pos = 4, key.width = lcm(5))
```

Comparision of spatial patterns for irregular local landscapes requires setting `window = ecoregions["id"]`.

```{r}
compare_2 = lsp_compare(landcover, landcoverold, 
                        type = "cove", dist_fun = "jensen-shannon", 
                        window = ecoregions["id"], threshold = 0.9)
```

Spatial patterns of land cover are very similar for most of the ecoregions, with just two exceptions:

```{r}
my_breaks = c(0, 0.001, 0.01, 0.1, 1.01)
tm_shape(compare_2) +
  tm_raster("dist", breaks = my_breaks, palette = "-viridis") +
  tm_shape(ecoregions) +
  tm_borders() +
  tm_layout(legend.outside = TRUE)
```

<!-- ADDS EXAMPLES -->
