---
title: "Types of spatial patterns' signatures"
author: "Jakub Nowosad"
date: "`r Sys.Date()`"
output: 
  html_document:
  github_document:
    html_preview: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

The fundamental element of the pattern-based spatial analysis is a numerical describtion of spatial patterns (so-called signature).
This vignette shows how to derive spatial signatures using the `lsp_signature()` function on example datasets.
Let's start by attaching neccesary packages:

```{r}
library(motif)
library(stars)
library(sf)
```

For this vignette, we use several spatial objects.
The first one is a `"raster/landcover2015.tif"` file.

```{r}
landcover = read_stars(system.file("raster/landcover2015.tif", package = "motif"))
```

This file contains a land cover data for New Guinea, with seven possible categories: (1) agriculture, (2) forest, (3) grassland, (5) settlement, (6) shrubland, (7) sparse vegetation, and (9) water.

```{r}
landcover = droplevels(landcover)
plot(landcover, key.pos = 4, key.width = lcm(5), main = NULL)
```

The second one is a `"raster/landform.tif"` file.

```{r}
landform = read_stars(system.file("raster/landform.tif", package = "motif"))
```

It has fourteen landform categories (plus surface water) for the same area.

```{r}
landform = droplevels(landform)
plot(landform, key.pos = 4, key.width = lcm(8), main = NULL)
```

The third example object is `random_ndvi`.

```{r}
set.seed(222)
random_ndvi = landcover
random_ndvi$ndvi = runif(length(random_ndvi[[1]]), min = 1, max = 10)
random_ndvi$ndvi[is.na(random_ndvi$landcover2015.tif)] = NA
random_ndvi$landcover2015.tif = NULL
```

It is an artificial dataset representing numerical weights for each cell for the same area as the datasets above.

```{r}
plot(random_ndvi)
```

## A co-occurrence matrix (`"coma"`) representation - one layer

The first type of signature is a co-occurrence matrix (`"coma"`).
It requires just one layer (one attribute in a `stars` object).

```{r}
coma_output = lsp_signature(landcover, type = "coma", window = 100)
coma_output
```

The output is an object of class `lsp` with three columns:

- `id` - an id of each window (area)
- `na_prop` - share (0-1) of `NA` cells for each window
- `signature` - a list-column containing with calculated signatures

We can see a signature for selected local landscape extracting it from the `signature` column: 

```{r}
coma_output$signature[[1]]
```

Co-occurrence vector (`"cove"`) is rearrangement of a co-occurrence matrix into one-dimensional object:

```{r}
cove_output = lsp_signature(landcover, type = "cove", window = 100)
cove_output
```

This representation can be used to compare different local landscapes.

Learn more about these representations at https://nowosad.github.io/comat/articles/coma.html.

## A weighted co-occurrence matrix (`wecoma`) representation - two layers

The next type of signature is a weighted co-occurrence matrix (`"wecoma"`).
It requires two layers - a `stars` object with two attributes.
The first one is a categorical raster data, while the second one is a continuous raster data containing weights.

```{r}
wecoma_output = lsp_signature(c(landcover, random_ndvi), 
                               type = "wecoma", window = 100)
wecoma_output
```

Weighted co-occurrence vector (`"wecove"`) is rearrangement of a weighted co-occurrence matrix into one-dimensional object:

```{r}
wecove_output = lsp_signature(c(landcover, random_ndvi),
                               type = "wecove", window = 100)
wecove_output
```

Learn more about these representations at https://nowosad.github.io/comat/articles/wecoma.html.

## An integrated co-occurrence matrix (`incoma`) representation - two or more layers

The next type of signature is an integrated co-occurrence matrix (`incoma`).
It requires two or more layers - a `stars` object with two or more attributes.
All layers must be categorical raster data.

```{r}
incoma_output = lsp_signature(c(landcover, landform),
                               type = "incoma", window = 100)
```

Integrated co-occurrence vector (`"incove"`) is rearrangement of an integrated co-occurrence matrix into one-dimensional object:

```{r}
incove_output = lsp_signature(c(landcover, landform),
                               type = "incove", window = 100)
```

Learn more about these representations at https://nowosad.github.io/comat/articles/incoma.html

## A composition representation (`"composition"`) - one layer

A composition signature describes proportions of categories in a local landscape.
It requires one layer (a `stars` object with one attribute).

```{r}
composition_output = lsp_signature(landcover, 
                                    type = "composition", window = 100)
composition_output
```

By default, it is normalized to sum to 1:

```{r}
composition_output$signature[[1]]
```

To get an actual number of cells of each category, the `normalization` should be set to `"none"`:

```{r}
composition_output2 = lsp_signature(landcover, 
                                    type = "composition", window = 100,
                                    normalization = "none")
composition_output2$signature[[1]]
```

## User-defined functions - one or more layers

The **motif** package also allows calculating signature based on a user-defined function.
This function should accept only one argument, which is a list containing one or more matrices.
For example `my_fun()` below counts how many non-NA cells exist in each local landscape.

```{r}
my_fun = function(x){
  sum(!is.na(c(x[[1]])))
} 
```

Importantly, we need to set `normalization = "none"` to prevent the normalization of the output.

```{r}
my_fun_output = lsp_signature(landcover, 
                               type = my_fun, window = 100,
                               normalization = "none")
my_fun_output
```

We can see that in the first local landscape we have `r my_fun_output$signature[[1]]` non-NA cells.

```{r}
my_fun_output$signature[[1]]
```

